---
title: "ANA515 Project Part 1 - R Script"
author: Anushree Das
output:
  pdf_document:
    keep_tex: true
---

```{r, message=FALSE}
#Load necessary libraries
library(dplyr)
library(lubridate)
library(tidyverse)
```

# 1. EXTRACT
```{r}
# Load dataset from csv
users_csv  <- read.csv("credit_card_transactions/sd254_users.csv")
cards_csv <- read.csv("credit_card_transactions/sd254_cards.csv")
transactions_csv <- read.csv("credit_card_transactions/credit_card_transactions-ibm_v2.csv")
```

# 2. TRANFORM
## 2.1 Modify users table
### 2.1.a Generate UserId column for users table
### 2.1.b Rename column names
```{r}
users <- data.frame(UserId = 1:nrow(users_csv), 
                        FullName = users_csv$Person, 
                        CurrentAge = users_csv$Current.Age,
                        RetirementAge = users_csv$Retirement.Age,
                        BirthYear = users_csv$Birth.Year, 
                        BirthMonth = users_csv$Birth.Month, 
                        Gender= users_csv$Gender, 
                        Address1 = users_csv$Address, 
                        Address2 = users_csv$Apartment, 
                        City = users_csv$City, 
                        State = users_csv$State, 
                        Zipcode = users_csv$Zipcode, 
                        Latitude = users_csv$Latitude, 
                        Longitude = users_csv$Longitude, 
                        PerCapitaIncomeZipcode = users_csv$Per.Capita.Income...Zipcode, 
                        YearlyIncome = users_csv$Yearly.Income...Person, 
                        TotalDebt = users_csv$Total.Debt, 
                        FICOScore = users_csv$FICO.Score, 
                        NumCreditCards = users_csv$Num.Credit.Cards
)

head(users, 3)
```

### 2.1.c Convert Gender to categorical data type
```{r}
users$Gender <- factor(users$Gender)
```

### 2.1.d Convert columns - Address2, Zipcode to Text data type
Zipcode and Address2 should be converted to Text data type because they are not used for quantitative statistics
```{r}
users$Address2 <- as.character(users$Address2)
users$Zipcode <- as.character(users$Zipcode)
```

### 2.1.e Convert columns - PerCapitaIncomeZipcode, YearlyIncome, TotalDebt to numeric
```{r}
# Remove $ character from the column values and convert to numeric data type
users$PerCapitaIncomeZipcode <- as.numeric(sub('.', '', users$PerCapitaIncomeZipcode))
users$YearlyIncome <- as.numeric(sub('.', '', users$YearlyIncome))
users$TotalDebt <- as.numeric(sub('.', '', users$TotalDebt))
```


## 2.2 Modify cards tables
### 2.2.a Remove duplicates
```{r}
cards_csv <- cards_csv %>% distinct() 
```

### 2.2.b Combine user id and card index to generate CardId 
### 2.2.c Rename column names
```{r}
cards <- data.frame(CardId = paste(cards_csv$User, cards_csv$CARD.INDEX, sep = "$"),
                        CardBrand = cards_csv$Card.Brand,
                        CardType = cards_csv$Card.Type,
                        CardNumber = cards_csv$Card.Number,
                        Expires = cards_csv$Expires,
                        CVV = cards_csv$CVV, 
                        HasChip = cards_csv$Has.Chip, 
                        CardsIssued = cards_csv$Cards.Issued, 
                        CreditLimit = cards_csv$Credit.Limit,
                        AcctOpenDate = cards_csv$Acct.Open.Date,
                        YearPINLastChanged = cards_csv$Year.PIN.last.Changed
)
head(cards, 3)
```

### 2.2.d Convert columns - CardBrand, CardType, HasChip to categorical data type
```{r}
cards$CardBrand <- factor(cards$CardBrand)
cards$CardType <- factor(cards$CardType)
cards$HasChip <- factor(cards$HasChip)
```

### 2.2.e Convert columns - Expires, AcctOpenDate to Date data type
The Date Format for Expires and AcctOpenDate contains only Month and Year, but R needs a value for Day as well.
The Expires date will be set to the last day of the month and AcctOpenDate to the first day of the month for this project.
```{r}
cards$Expires <- my(cards$Expires)
cards$Expires <- ceiling_date(cards$Expires, "month") - days(1)

cards$AcctOpenDate <- my(cards$AcctOpenDate)
```

### 2.2.f Convert columns - CardNumber, CVV to Text data type
CardNumber and CVV not used for quantitative statistics, therefore they are being converted to Text 
```{r}
cards$CardNumber <- as.character(cards$CardNumber)
cards$CVV <- as.character(cards$CVV)
```

### 2.2.g Convert column CreditLimit to numeric
```{r}
cards$CreditLimit <- as.numeric(sub('.', '', cards$CreditLimit))
```

## 2.3 Create table for merchants
### 2.3.a Extract Merchant related column from the Transactions table
```{r}
merchants_temp <- data.frame(MerchantName = transactions_csv$Merchant.Name,
                            MerchantCity = transactions_csv$Merchant.City,
                            MerchantState = transactions_csv$Merchant.State,
                            MerchantZIP = transactions_csv$Zip,
                            MerchantCategoryCode = transactions_csv$MCC
)
```

### 2.3.b Remove redundant data from the Merchants table
```{r}
nrow(merchants_temp)
length(unique(merchants_temp$MerchantName))
merchants_temp <- merchants_temp %>% distinct(MerchantName, .keep_all = TRUE)
nrow(merchants_temp)
```

### 2.3.c Generate and add MerchantId column
```{r}
merchants <- data.frame(MerchantId = 1:nrow(merchants_temp),
                            MerchantName = merchants_temp$MerchantName,
                            MerchantCity = merchants_temp$MerchantCity,
                            MerchantState = merchants_temp$MerchantState,
                            MerchantZIP = merchants_temp$MerchantZIP,
                            MerchantCategoryCode = merchants_temp$MerchantCategoryCode
)

head(merchants, 3)
```

### 2.3.d Convert columns - MerchantName, MerchantZIP to Text data type
```{r}
merchants$MerchantName <- as.character(merchants$MerchantName)
merchants$MerchantZIP <- as.character(merchants$MerchantZIP)
```

### 2.3.e Convert column MerchantCategoryCode to Categorical data type
```{r}
merchants$MerchantCategoryCode = factor(merchants$MerchantCategoryCode)
```


## 2.4 Create table for date
### 2.4.a Transform Time column to categorical (morning, afternoon, evening, night)
```{r}
# Convert column to Time data type
transactions_csv$Time <- as.POSIXct(strptime(transactions_csv$Time,"%H:%M"),"UTC")

# Define time range for each category
x=as.POSIXct(strptime(c("050000","105959","110000","155959","160000",
                        "185959"),"%H%M%S"),"UTC")

# Transform Time column 
transactions_csv$Time = case_when(
  between(transactions_csv$Time,x[1],x[2]) ~"morning",
  between(transactions_csv$Time,x[3],x[4]) ~"afternoon",
  between(transactions_csv$Time,x[5],x[6]) ~"evening",
  TRUE ~"night")

head(transactions_csv$Time)
```

### 2.4.b Extract Date related column from the Transactions table
```{r}
date_temp <- data.frame(DateYear = transactions_csv$Year,
                        DateMonth = transactions_csv$Month,
                        DateDay = transactions_csv$Day,
                        Time = transactions_csv$Time
)
```

### 2.4.c Remove redundant data from the Date table
```{r}
date_temp <- date_temp %>% distinct() 
```

### 2.4.d Generate and add DateId column
```{r}
date <- data.frame(DateId = 1:nrow(date_temp),
                        DateYear = date_temp$DateYear,
                        DateMonth = date_temp$DateMonth,
                        DateDay = date_temp$DateDay,
                        Time = date_temp$Time
)
```

### 2.4.e Convert column Time to Categorical data type
```{r}
date$Time = factor(date$Time)
```


## 2.5 Fact table - transactions
### 2.4.a Replace existing merchant columns with single merchant id
```{r, message=FALSE, warning = FALSE}
# Uses Merchant Name (unique column in merchants table) 
# to assign Merchant Id values in Transactions table
transactions_csv$MerchantId <- transactions_csv$Merchant.Name
conditions <- merchants$MerchantName
replacement_values <- merchants$MerchantId
transactions_csv$MerchantId <- replace(transactions_csv$MerchantId, 
                                   transactions_csv$MerchantId %in% conditions, 
                                   replacement_values)
```

### 2.4.b Replace existing date columns with single date id
```{r, message=FALSE, warning = FALSE}
# Combine all columns in Dates table
temp_dates = data.frame(DateId = date$DateId,
                        Condition = paste(date$DateYear,
                                          date$DateMonth,
                                          date$DateDay,
                                          date$Time)
)

# Combine all columns that were added to Date table from Transaction table 
transactions_csv$Condition <- paste(transactions_csv$Year,
                                transactions_csv$Month,
                                transactions_csv$Day,
                                transactions_csv$Time)

# Add DateId column with temporary values
transactions_csv$DateId <- 1:nrow(transactions_csv)

# Replace DateId based on the combined Date columns in both tables
transactions_csv$DateId <- replace(transactions_csv$DateID, 
                             transactions_csv$Condition %in% temp_dates$Condition, 
                             temp_dates$DateId)
```

### 2.4.c Remove duplicates
```{r}
transactions_csv <- transactions_csv %>% distinct() 
```

### 2.4.d Combine user id and card index to generate CardId column 
### 2.4.e Rename column names
```{r}
transactions <- data.frame(TransactionId = 1:nrow(transactions_csv),
                              CardId = paste(transactions_csv$User, 
                                             transactions_csv$Card, sep = "$"),
                              UserId = transactions_csv$User,
                              DateId = transactions_csv$DateId,
                              MerchantId = transactions_csv$MerchantId,
                              Amount = transactions_csv$Amount, 
                              UseChip = transactions_csv$Use.Chip, 
                              Errors = transactions_csv$Errors., 
                              IsFraud = transactions_csv$Is.Fraud
)

head(transactions, 3)
```

### 2.4.f Covert columns - UseChip, IsFraud to categorical data type
```{r}
transactions$UseChip <- factor(transactions$UseChip)
transactions$IsFraud <- factor(transactions$IsFraud)
```

### 2.2.g Convert column Amount to numeric
```{r}
transactions$Amount <- as.numeric(sub('.', '', transactions$Amount))
```

# 3. LOAD
## Dimension table - users_dim
```{r}
users_dim <- users %>% select(UserId, FullName, CurrentAge, RetirementAge, 
                              BirthYear, BirthMonth, Gender, Address1, Address2, 
                              City, State, Zipcode, Latitude, Longitude, 
                              PerCapitaIncomeZipcode, YearlyIncome, TotalDebt, 
                              FICOScore, NumCreditCards) %>% distinct()
summary(users_dim)
```

## Dimension table - cards_dim
```{r}
cards_dim <- cards %>% select(CardId, CardBrand, CardType, CardNumber, Expires, 
                              CVV, HasChip, CardsIssued, CreditLimit, 
                              AcctOpenDate, YearPINLastChanged) %>% distinct()
summary(cards)
```

## Dimension table - cards_dim
```{r}
merchants_dim <- merchants %>% select(MerchantId, MerchantName, MerchantCity, 
                                      MerchantState, MerchantZIP, 
                                      MerchantCategoryCode) %>% distinct()
summary(merchants_dim)
```

## Dimension table - date_dim
```{r}
date_dim <- date %>% select(DateId, DateYear, DateMonth, 
                            DateDay, Time) %>% distinct()
summary(date_dim)
```

## Fact table - transactions_fact
```{r}
transactions_fact <- transactions %>% select(TransactionId, UserId, CardId, 
                                             DateId, Amount, UseChip, Errors, 
                                             IsFraud) %>% distinct()
summary(transactions_fact)
```
