---
title: "ANA515 Project Part 2 - R Script"
author: Anushree Das
output:
  pdf_document:
    keep_tex: true
---

For new changes added to the pipeline, jump to [Advanced Dimensional Modeling](#adv-modeling)
```{r, message=FALSE}
#Load necessary libraries
library(dplyr)
library(lubridate)
library(tidyverse)
library(ggplot2)
```

# 1. EXTRACT
```{r}
# Load dataset from csv
users_csv  <- read.csv("credit_card_transactions/sd254_users.csv")
cards_csv <- read.csv("credit_card_transactions/sd254_cards.csv")
transactions_csv <- read.csv("credit_card_transactions/credit_card_transactions-ibm_v2.csv")
```

# 2. TRANFORM
## 2.1 Modify users table
### 2.1.a Generate UserId column for users table
### 2.1.b Rename column names
```{r}
users <- data.frame(UserId = 1:nrow(users_csv), 
                        FullName = users_csv$Person, 
                        CurrentAge = users_csv$Current.Age,
                        RetirementAge = users_csv$Retirement.Age,
                        BirthYear = users_csv$Birth.Year, 
                        BirthMonth = users_csv$Birth.Month, 
                        Gender= users_csv$Gender, 
                        Address1 = users_csv$Address, 
                        Address2 = users_csv$Apartment, 
                        City = users_csv$City, 
                        State = users_csv$State, 
                        Zipcode = users_csv$Zipcode, 
                        Latitude = users_csv$Latitude, 
                        Longitude = users_csv$Longitude, 
                        PerCapitaIncomeZipcode = users_csv$Per.Capita.Income...Zipcode, 
                        YearlyIncome = users_csv$Yearly.Income...Person, 
                        TotalDebt = users_csv$Total.Debt, 
                        FICOScore = users_csv$FICO.Score, 
                        NumCreditCards = users_csv$Num.Credit.Cards
)

head(users, 3)
```

### 2.1.c Convert Gender to categorical data type
```{r}
users$Gender <- factor(users$Gender)
```

### 2.1.d Convert columns - Address2, Zipcode to Text data type
Zipcode and Address2 should be converted to Text data type because they are not used for quantitative statistics
```{r}
users$Address2 <- as.character(users$Address2)
users$Zipcode <- as.character(users$Zipcode)
```

### 2.1.e Convert columns - PerCapitaIncomeZipcode, YearlyIncome, TotalDebt to numeric
```{r}
# Remove $ character from the column values and convert to numeric data type
users$PerCapitaIncomeZipcode <- as.numeric(sub('.', '', users$PerCapitaIncomeZipcode))
users$YearlyIncome <- as.numeric(sub('.', '', users$YearlyIncome))
users$TotalDebt <- as.numeric(sub('.', '', users$TotalDebt))
```


## 2.2 Modify cards tables
### 2.2.a Remove duplicates
```{r}
cards_csv <- cards_csv %>% distinct() 
```

### 2.2.b Combine user id and card index to generate CardId 
### 2.2.c Rename column names
```{r}
cards <- data.frame(CardId = paste(cards_csv$User, cards_csv$CARD.INDEX, sep = "$"),
                        CardBrand = cards_csv$Card.Brand,
                        CardType = cards_csv$Card.Type,
                        CardNumber = cards_csv$Card.Number,
                        Expires = cards_csv$Expires,
                        CVV = cards_csv$CVV, 
                        HasChip = cards_csv$Has.Chip, 
                        CardsIssued = cards_csv$Cards.Issued, 
                        CreditLimit = cards_csv$Credit.Limit,
                        AcctOpenDate = cards_csv$Acct.Open.Date,
                        YearPINLastChanged = cards_csv$Year.PIN.last.Changed
)
head(cards, 3)
```

### 2.2.d Convert columns - CardBrand, CardType, HasChip to categorical data type
```{r}
cards$CardBrand <- factor(cards$CardBrand)
cards$CardType <- factor(cards$CardType)
cards$HasChip <- factor(cards$HasChip)
```

### 2.2.e Convert columns - Expires, AcctOpenDate to Date data type
The Date Format for Expires and AcctOpenDate contains only Month and Year, but R needs a value for Day as well.
The Expires date will be set to the last day of the month and AcctOpenDate to the first day of the month for this project.
```{r}
cards$Expires <- my(cards$Expires)
cards$Expires <- ceiling_date(cards$Expires, "month") - days(1)

cards$AcctOpenDate <- my(cards$AcctOpenDate)
```

### 2.2.f Convert columns - CardNumber, CVV to Text data type
CardNumber and CVV not used for quantitative statistics, therefore they are being converted to Text 
```{r}
cards$CardNumber <- as.character(cards$CardNumber)
cards$CVV <- as.character(cards$CVV)
```

### 2.2.g Convert column CreditLimit to numeric
```{r}
cards$CreditLimit <- as.numeric(sub('.', '', cards$CreditLimit))
```

## 2.3 Create table for merchants
### 2.3.a Extract Merchant related column from the Transactions table
```{r}
merchants_temp <- data.frame(MerchantName = transactions_csv$Merchant.Name,
                            MerchantCity = transactions_csv$Merchant.City,
                            MerchantState = transactions_csv$Merchant.State,
                            MerchantZIP = transactions_csv$Zip,
                            MerchantCategoryCode = transactions_csv$MCC
)
```

### 2.3.b Remove redundant data from the Merchants table
```{r}
nrow(merchants_temp)
length(unique(merchants_temp$MerchantName))
merchants_temp <- merchants_temp %>% distinct(MerchantName, .keep_all = TRUE)
nrow(merchants_temp)
```

### 2.3.c Generate and add MerchantId column
```{r}
merchants <- data.frame(MerchantId = 1:nrow(merchants_temp),
                            MerchantName = merchants_temp$MerchantName,
                            MerchantCity = merchants_temp$MerchantCity,
                            MerchantState = merchants_temp$MerchantState,
                            MerchantZIP = merchants_temp$MerchantZIP,
                            MerchantCategoryCode = merchants_temp$MerchantCategoryCode
)

head(merchants, 3)
```

### 2.3.d Convert columns - MerchantName, MerchantZIP to Text data type
```{r}
merchants$MerchantName <- as.character(merchants$MerchantName)
merchants$MerchantZIP <- as.character(merchants$MerchantZIP)
```

### 2.3.e Convert column MerchantCategoryCode to Categorical data type
```{r}
merchants$MerchantCategoryCode = factor(merchants$MerchantCategoryCode)
```


## 2.4 Create table for date
### 2.4.a Transform Time column to categorical (morning, afternoon, evening, night)
```{r}
# Convert column to Time data type
transactions_csv$Time <- as.POSIXct(strptime(transactions_csv$Time,"%H:%M"),"UTC")

# Define time range for each category
x=as.POSIXct(strptime(c("050000","105959","110000","155959","160000",
                        "185959"),"%H%M%S"),"UTC")

# Transform Time column 
transactions_csv$Time = case_when(
  between(transactions_csv$Time,x[1],x[2]) ~"morning",
  between(transactions_csv$Time,x[3],x[4]) ~"afternoon",
  between(transactions_csv$Time,x[5],x[6]) ~"evening",
  TRUE ~"night")

head(transactions_csv$Time)
```

### 2.4.b Extract Date related column from the Transactions table
```{r}
date_temp <- data.frame(DateYear = transactions_csv$Year,
                        DateMonth = transactions_csv$Month,
                        DateDay = transactions_csv$Day,
                        Time = transactions_csv$Time
)
```

### 2.4.c Remove redundant data from the Date table
```{r}
date_temp <- date_temp %>% distinct() 
```

### 2.4.d Generate and add DateId column
```{r}
date <- data.frame(DateId = 1:nrow(date_temp),
                        DateYear = date_temp$DateYear,
                        DateMonth = date_temp$DateMonth,
                        DateDay = date_temp$DateDay,
                        Time = date_temp$Time
)
```

### 2.4.e Convert column Time to Categorical data type
```{r}
date$Time = factor(date$Time)
```

## 2.5 Fact table - transactions
### 2.5.a Replace existing merchant columns with single merchant id
```{r, message=FALSE, warning = FALSE}
# Uses Merchant Name (unique column in merchants table) 
# to assign Merchant Id values in Transactions table
transactions_csv$MerchantId <- transactions_csv$Merchant.Name
conditions <- merchants$MerchantName
replacement_values <- merchants$MerchantId
transactions_csv$MerchantId <- replace(transactions_csv$MerchantId, 
                                   transactions_csv$MerchantId %in% conditions, 
                                   replacement_values)
```

### 2.5.b Replace existing date columns with single date id
```{r, message=FALSE, warning = FALSE}
# Combine all columns in Dates table
temp_dates = data.frame(DateId = date$DateId,
                        Condition = paste(date$DateYear,
                                          date$DateMonth,
                                          date$DateDay,
                                          date$Time)
)

# Combine all columns that were added to Date table from Transaction table 
transactions_csv$Condition <- paste(transactions_csv$Year,
                                transactions_csv$Month,
                                transactions_csv$Day,
                                transactions_csv$Time)

# Add DateId column with temporary values
transactions_csv$DateId <- 1:nrow(transactions_csv)

# Replace DateId based on the combined Date columns in both tables
transactions_csv$DateId <- replace(transactions_csv$DateID, 
                             transactions_csv$Condition %in% temp_dates$Condition, 
                             temp_dates$DateId)
```

### 2.5.c Remove duplicates
```{r}
transactions_csv <- transactions_csv %>% distinct() 
```

### 2.5.d Add TransactionId column
```{r}
transactions_csv$TransactionId <- 1:nrow(transactions_csv)
```

### 2.5.e Combine user id and card index to generate CardId column 
### 2.5.f Rename column names
```{r}
transactions <- data.frame(TransactionId = transactions_csv$TransactionId,
                              CardId = paste(transactions_csv$User, 
                                             transactions_csv$Card, sep = "$"),
                              UserId = transactions_csv$User,
                              DateId = transactions_csv$DateId,
                              MerchantId = transactions_csv$MerchantId,
                              Amount = transactions_csv$Amount
)

head(transactions, 3)
```

### 2.5.g Convert column Amount to numeric
```{r}
transactions$Amount <- as.numeric(sub('.', '', transactions$Amount))
```

# 3. Advanced Dimensional Modeling {#adv-modeling}
## 3.1 Junk Dimension table - transaction_flags
### 3.1.a Create table
```{r}
transaction_flags <- data.frame(TransactionId = transactions_csv$TransactionId,
                                UseChip = transactions_csv$Use.Chip, 
                                IsFraud = transactions_csv$Is.Fraud
)
head(transaction_flags, 3)
```

### 3.1.b Covert columns - UseChip, IsFraud to categorical data type
```{r}
transaction_flags$UseChip <- factor(transaction_flags$UseChip)
transaction_flags$IsFraud <- factor(transaction_flags$IsFraud)
```

## 3.2 Bridge Table
The Transaction and Error Types have many-to-many relationship.

### 3.2.a Create bridge table by splitting multiple errors
```{r}
nrow(transactions_csv)
# Split the errors list for each transaction
transaction_error_bridge <- transactions_csv %>% separate_longer_delim(Errors., delim = ",")
# Rename column
transaction_error_bridge$ErrorDescription <- transaction_error_bridge$Errors.
nrow(transaction_error_bridge)
```

### 3.2.b Create error dimension table with unique ErrorId
```{r}
errors <- data.frame(ErrorDescription=distinct(transaction_error_bridge,ErrorDescription))
# remove Error Description with empty string
errors <- errors %>% filter(ErrorDescription != "")
# add id column
errors$ErrorId <- 1:length(errors$ErrorDescription)
head(errors)
```

### 3.2.c Add ErrorId to bridge table
```{r}
transaction_error_bridge <- transaction_error_bridge %>%
  left_join(errors, by = "ErrorDescription") %>%
  select(TransactionId, ErrorId)
```

# 4. Data Quality Assessment
## 4.1 Identifying and Handling Missing values
### Identify missing values in users table
```{r}
colSums(is.na(users))
```
Address2 has 1472 NA values

### Identify missing values in cards table
```{r}
colSums(is.na(cards))
```

### Identify missing values in merchants table
```{r}
colSums(is.na(merchants))
```
MerchantZIP has 1267 NA values

### Identify missing values in date table
```{r}
colSums(is.na(date))
```

### Identify missing values in transactions table
```{r}
colSums(is.na(transactions))
```

### Identify missing values in transaction_flags table
```{r}
colSums(is.na(transaction_flags))
```
### Handle missing values 
1. Replace Address2 column with "" string, since it is usually an optional variable
```{r}
users$Address2[is.na(users$Address2)] <- ""
```
2. Drop rows where ZIP is NA, since it is an important feature for location based analysis of merchants and the count of rows with missing zip is very small(1267) when compared with the total number of rows(100,343) in the merchants table
```{r}
merchants <- na.omit(merchants)
```

## 4.2 Detecting and Handling Outliers
### 4.2.a Detect outliers in Amount column in Transactions table
```{r}
ggplot(transactions, aes(x=Amount)) + geom_density()
```
```{r}
summary(transactions$Amount)
```

### Handle outliers
Since, the number of outliers (Amount>500) is large, setting them to the value of upper bound calculated using IQR method would introduce skewness.

Better approach would be to transform the column. Group outliers into categories "Low", "Medium", "High", "Extreme"
```{r}
transactions <- transactions %>%
  mutate(AmountCategory = case_when(
    Amount <= 150 ~ "Low",
    Amount <= 500 ~ "Medium",
    Amount <= 2000 ~ "High",
    TRUE ~ "Extreme"
  ))
```

### 4.2.b Detect outliers in CurrentAge column in Users table
```{r}
Q1 <- quantile(users$CurrentAge, 0.25)
Q3 <- quantile(users$CurrentAge, 0.75)
IQR <- Q3 - Q1
# Calculate upper and lower bound
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Identify outliers
boxplot.stats(users$CurrentAge)$out
```
### Handle ouliers
```{r}
users$CurrentAge[users$CurrentAge < lower_bound] <- lower_bound
users$CurrentAge[users$CurrentAge > upper_bound] <- upper_bound
```

### 4.2.c Detect outliers in RetirementAge column in Users table
```{r}
Q1 <- quantile(users$RetirementAge, 0.25)
Q3 <- quantile(users$RetirementAge, 0.75)
IQR <- Q3 - Q1
# Calculate upper and lower bound
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Identify outliers
boxplot.stats(users$RetirementAge)$out
```
### Handle ouliers
```{r}
users$RetirementAge[users$RetirementAge < lower_bound] <- lower_bound
users$RetirementAge[users$RetirementAge > upper_bound] <- upper_bound
```

### 4.2.d Detect outliers in PerCapitaIncomeZipcode column in Users table
```{r}
Q1 <- quantile(users$PerCapitaIncomeZipcode, 0.25)
Q3 <- quantile(users$PerCapitaIncomeZipcode, 0.75)
IQR <- Q3 - Q1
# Calculate upper and lower bound
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Identify outliers
boxplot.stats(users$PerCapitaIncomeZipcode)$out
```
### Handle ouliers
```{r}
users$PerCapitaIncomeZipcode[users$PerCapitaIncomeZipcode < lower_bound] <- lower_bound
users$PerCapitaIncomeZipcode[users$PerCapitaIncomeZipcode > upper_bound] <- upper_bound
```

### 4.2.e Detect outliers in YearlyIncome column in Users table
```{r}
Q1 <- quantile(users$YearlyIncome, 0.25)
Q3 <- quantile(users$YearlyIncome, 0.75)
IQR <- Q3 - Q1
# Calculate upper and lower bound
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Identify outliers
boxplot.stats(users$YearlyIncome)$out
```
### Handle ouliers
```{r}
users$YearlyIncome[users$YearlyIncome < lower_bound] <- lower_bound
users$YearlyIncome[users$YearlyIncome > upper_bound] <- upper_bound
```

### 4.2.f Detect outliers in TotalDebt column in Users table
```{r}
Q1 <- quantile(users$TotalDebt, 0.25)
Q3 <- quantile(users$TotalDebt, 0.75)
IQR <- Q3 - Q1
# Calculate upper and lower bound
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Identify outliers
boxplot.stats(users$TotalDebt)$out
```
### Handle ouliers
```{r}
users$TotalDebt[users$TotalDebt < lower_bound] <- lower_bound
users$TotalDebt[users$TotalDebt > upper_bound] <- upper_bound
```

### 4.2.g Detect outliers in CreditLimit column in cards table
```{r}
Q1 <- quantile(cards$CreditLimit, 0.25)
Q3 <- quantile(cards$CreditLimit, 0.75)
IQR <- Q3 - Q1
# Calculate upper and lower bound
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Identify outliers
boxplot.stats(cards$CreditLimit)$out
```
### Handle ouliers
```{r}
cards$CreditLimit[cards$CreditLimit < lower_bound] <- lower_bound
cards$CreditLimit[cards$CreditLimit > upper_bound] <- upper_bound
```

# 5. Custom Functions in R
## 5.1 Standardize numeric variables to ensure consistent range across features
```{r}
users$CurrentAgeScaled <- scale(users$CurrentAge)
users$RetirementAgeScaled <- scale(users$RetirementAge)
users$PerCapitaIncomeZipcodeScaled <- scale(users$PerCapitaIncomeZipcode)
users$YearlyIncomeScaled <- scale(users$YearlyIncome)
users$TotalDebtScaled <- scale(users$TotalDebt)
users$FICOScoreScaled <- scale(users$FICOScore)

cards$CreditLimitScaled <- scale(cards$CreditLimit)
```

## 5.2 Convert text columns to lowercase and trim whitespace 
```{r}
users <- users %>%  mutate(across(where(is.character), str_trim))
users <- users %>%  mutate(across(where(is.character), tolower))  

merchants <- merchants %>%  mutate(across(where(is.character), str_trim))
merchants <- merchants %>%  mutate(across(where(is.character), tolower))  
```

# 6. LOAD
## Dimension table - users_dim
```{r}
users_dim <- users %>% select(UserId, FullName, CurrentAgeScaled, RetirementAgeScaled, 
                              BirthYear, BirthMonth, Gender, Address1, Address2, 
                              City, State, Zipcode, Latitude, Longitude, 
                              PerCapitaIncomeZipcodeScaled, YearlyIncomeScaled, TotalDebtScaled, 
                              FICOScoreScaled, NumCreditCards) %>% distinct()
summary(users_dim)
```
```{r}
head(users_dim,4)
```

## Dimension table - cards_dim
```{r}
cards_dim <- cards %>% select(CardId, CardBrand, CardType, CardNumber, Expires, 
                              CVV, HasChip, CardsIssued, CreditLimitScaled, 
                              AcctOpenDate, YearPINLastChanged) %>% distinct()
summary(cards_dim)
```
```{r}
head(cards_dim,4)
```

## Dimension table - merchants_dim
```{r}
merchants_dim <- merchants %>% select(MerchantId, MerchantName, MerchantCity, 
                                      MerchantState, MerchantZIP, 
                                      MerchantCategoryCode) %>% distinct()
summary(merchants_dim)
```
```{r}
head(merchants_dim,4)
```
## Dimension table - date_dim
```{r}
date_dim <- date %>% select(DateId, DateYear, DateMonth, 
                            DateDay, Time) %>% distinct()
summary(date_dim)
```

## Dimension table - errors_dim
```{r}
errors_dim <- errors %>% select(ErrorId, ErrorDescription) %>% distinct()
summary(errors_dim)
```

## Bridge table - transaction_error_bridge
```{r}
transaction_error_bridge <- transaction_error_bridge %>% 
  select(TransactionId, ErrorId) %>% distinct()
summary(transaction_error_bridge)
```

## Junk Dimension table - transaction_flags
```{r}
transaction_flags <- transaction_flags %>% select(TransactionId, UseChip, IsFraud)   
summary(transaction_flags)
```

## Fact table - transactions_fact
```{r}
transactions_fact <- transactions %>% select(TransactionId, UserId, CardId, 
                                             DateId, AmountCategory
                                             ) %>% distinct()
summary(transactions_fact)
```